Repeated loss of variation in insect ovary morphology highlights the role of developmental constraint in life-history evolution - Supplementary Methods

### 2021-01-18 manuscript_supplement.Rmd
adding conda environment; rerunning analyses; updating manuscript and README

---
title: Repeated loss of variation in insect ovary morphology highlights the role of developmental constraint in life-history evolution - Supplementary Methods
output:
    bookdown::pdf_document2:
    keep_tex: yes
    latex_engine: pdflatex
    #bookdown::html_document2: default
toc: true
bibliography: manuscript_supplement.bib
csl: nature.csl
link-citations: yes
header-includes:
 \usepackage{float}
 \usepackage{array}
 \usepackage{xcolor,colortbl}
 \floatplacement{figure}{H}
---

@@

```{r load_data}
# This script reads in and formats the ovariole number dataset
source("analysis/get_ovariole_number_data.R")
```

Samuel H. Church^1^\*, Bruno A. S. De Medeiros^1,2^, Seth Donoughe^1,3^, Nicole L. Márquez Reyes^4^, Cassandra G. Extavour^1,5^\*

@@

Ovariole number was recorded as either an average with deviations, a range, or a single total value, with priority for recording given in that order when multiple data types were available. Ovariole number was recorded as the total number of ovarioles per female, summing over both the left and right adult ovaries. When authors reported ovariole number from a single ovary, the total value was calculated by doubling the reported value. When authors described differences between the two ovaries, this information was recorded in an additional column.

@@

We combined the data we collected on total ovariole number with existing datasets of egg size and shape[@church2019dataset], insect lifetime fecundity and adult dry body weight[@Gilbert2007], average adult body length per insect family[@rainford2016phylogenetic], and several lineage-specific measures of adult body size[@iwata1966large;@Iwata1966a;@waloff1954number;@starmer2003phylogenetic;@reinhardt2005low].

Ovariole number and egg size[@church2019insect] data were combined by matching records at the species level (Fig 2a). When multiple records existed for a given species, the dataset was randomly shuffled and a single matching record was selected. This variation across records for the same species was accounted for in regressions by reshuffling and matching records at each iteration of the analysis. In order to test whether results were robust with a larger sample size, we also matched records at the genus level following the same reshuffling method (Fig. \ref{fig:PGLS-egg-ON-genus}).

@@

Ovariole number, egg volume, lifetime fecundity, and adult body weight[@Gilbert2007] were combined by matching records at the species level and genus level, using the same method as described above (Figs. 2b, 3, and \ref{fig:PGLS-Gilbert-name}). We excluded one value from this dataset which appeared to include a typographical error for lifetime fecundity (Hymenoptera: Trichogrammatidae, lifetime fecundity recorded as 0.1[@Gilbert2007]).

@@

set.seed(12345)  # set seed for cobmination
source("analysis/combine_body_egg_datasets.R")

@@

```{r PGLS-Gilbert-name, fig.cap="**Egg volume vs ovariole number, phylogenetic residuals to dry adult body weight, matching records at the species level.**", fig.width=3, fig.height=3, include=TRUE}
### Gilbert egg vs ON, ind body weight name pgls
egg_ON_gilbert_name_pgls <- egg_ON_gilbert_name %>% 
    select(genus,logvol,logon,logweight,group) %>% 
    mutate(rank = genus, trait1 = logon, trait2= logvol, indep = logweight)

@@

\textcolor{red}{We ran each Phylogenetic Generalized Least Squares regression over the Maximum Clade Credibilty (MCC) tree. For each regression, we also repeated each PGLS analysis 1000 times,} accounting for phylogenetic and phenotypic uncertainty, using the R packages ape (version `r packageVersion('ape')`)[@paradis2004ape] and nlme (version `r packageVersion('nlme')`)[@pinheiro2014r]. In these analyses we used a Brownian Motion based covariance matrix for traits.

@@

\textcolor{red}{For regressions of egg size and ovariole number when accounting for adult body size, we compared the results of our regression analyses to distributions estimated using simulated data under alternative hypotheses. We fit a Brownian motion model to the phylogenetic residuals of egg size and body size (R package geiger}[@harmon2007geiger]\textcolor{red}{, version `r packageVersion('geiger')`), and then used the parameters of this fitted model to simulate new datasets (R package phylolm}[@tung2014linear]\textcolor{red}{, version `r packageVersion('phylolm')`). We performed this resimulation using the datasets of egg size and body length at the family level, and egg size and body weight at the genus level. We simulated 1000 datasets each under two hypotheses: no correlation (slope=0) and a strong negative correlation (slope=-1:1). We performed the regressions as described above and compared the distribution of p-values and slopes to values from regressions on observed data.}

```{r simulated, fig.cap="**\\textcolor{red}{Using simulated data to test alternative hypotheses of evolutionary relationships}** Top row, distributions of p-values over 1000 replicate regressions, dashed black line indicates threshold of 0.01. Bottom row, distribution of estimated slopes between egg size and ovariole number, dashed lines indicate 95% interval of simulated distributions. Left, comparing egg size and ovariole number, accounting for body length at the family level. Right, comparing egg size and ovariole number, accounting for body weight at the genus level. Red=observed values, gray=simulated with no correlation, black=simulated with a -1:1 correlation. n=1000 regressions.", fig.width=7.5, fig.height=5, include=TRUE}
get_tstat <- function(sum,n) {
    tstat <- as.numeric(unname(sum[[n]]$coefficients[[2]]))
    return(tstat)
}

@@

load("analysis/pgls_results_Fall2020/run_pgls_simulated_allometry_body_egg_ON_family_-1_image.RData")
sim_corr_tstat_dist <- sapply(allometry_distribution_raw,get_tstat,n=1)
sim_corr_pval_dist <- sapply(allometry_distribution_raw,get_pval,n=1)
load("analysis/pgls_results_Fall2020/run_pgls_simulated_allometry_body_egg_ON_family_0_image.RData")
sim_nocorr_tstat_dist <- sapply(allometry_distribution_raw,get_tstat,n=1)
sim_nocorr_pval_dist <- sapply(allometry_distribution_raw,get_pval,n=1)
load("analysis/pgls_results_Fall2020/run_posterior_pgls_egg_size_ovariole_number.RData")
obs_tstat_dist <- sapply(allometry_distribution_raw,get_tstat,n=3)
obs_pval_dist <- sapply(allometry_distribution_raw,get_pval,n=3)

@@

load("analysis/pgls_results_Fall2020/run_pgls_simulated_allometry_gilbert_genus_-1_image.RData")
sim_corr_tstat_dist <- sapply(allometry_distribution_raw,get_tstat,n=1)
sim_corr_pval_dist <- sapply(allometry_distribution_raw,get_pval,n=1)
load("analysis/pgls_results_Fall2020/run_pgls_simulated_allometry_gilbert_genus_0_image.RData")
sim_nocorr_tstat_dist <- sapply(allometry_distribution_raw,get_tstat,n=1)
sim_nocorr_pval_dist <- sapply(allometry_distribution_raw,get_pval,n=1)
load("analysis/pgls_results_Fall2020/run_posterior_pgls_egg_size_ovariole_number.RData")
obs_tstat_dist <- sapply(allometry_distribution_raw,get_tstat,n=5)
obs_pval_dist <- sapply(allometry_distribution_raw,get_pval,n=5)

@@

p2 <- ggplot(gilbert_genus_pval_dist,aes(x=pval,y=relationship,color=relationship)) + geom_jitter(size=0.2)  + 
geom_vline(xintercept=0.01,colour="black",linetype="dashed") + 
#geom_vline(xintercept=0.05,colour="black",linetype="longdash") + 
ggtitle("egg size vs ov. number, body weight") + 
xlab("p-value") +
#ylab("type") + 
xlim(c(0,1)) +
scale_color_manual(values = c("black","dark gray","red")) +
theme(legend.position="none",
    axis.title.y=element_blank(),
axis.text.y=element_blank(),
axis.ticks.y=element_blank())

@@

```{r PGLS-body-plot, fig.cap="**Adult body size vs ovariole number.** Adult body weight compared at the genus level (left), and average adult body length compared at the family level (right).", fig.width=6, fig.height=3, include=TRUE}
### body length vs ON, family level
ON_body_family_pgls <- body_egg_ON_full %>% 
    select(rank,logon,logbody,group) %>% 
    rename(trait1 = logon, trait2 = logbody)

@@

gilbert_body <- run_plotting_pgls(ON_body_gilbert_genus_pgls,tree,"body_gilbert") + xlab(bquote('ovariole number,'~log[10])) +
ylab(bquote('adult dry body weight (g),'~log[10]))

@@

### PGLS TABLE EGG SIZE OVARIOLE NUMBER
load("analysis/pgls_results_Fall2020/run_posterior_pgls_egg_size_ovariole_number.RData")

@@

egg_ON_pgls_analysis_names <- c("ovariole number vs egg volume",
    "ovariole number vs egg volume",
    "ovariole number vs egg volume, residuals to body weight",
    "ovariole number vs egg volume, residuals to body weight",
    "ovariole number vs egg volume, residuals to body length")

@@

egg_ON_table <- as.data.frame(bind_rows(lapply(egg_ON_posterior_results,get_pgls_table))) %>% mutate(analysis = egg_ON_pgls_analysis_names,
        "taxonomic level" = egg_ON_taxonomic_levels,
        "slope" = paste(sprintf("%.3f",slope_min),sprintf("%.3f",slope_max),sep=sp),
        "MCC p-value" = gsub("0.000","<0.001",sprintf("%.3f",sapply(egg_ON_pgls_results,function(x){x$tTable[[8]]}))),
        "num. sig. / 1000" = pval_sig,
        "taxa" = taxa)
knitr::kable(egg_ON_table %>% select(c("analysis","taxonomic level","slope","MCC p-value","num. sig. / 1000","taxa")),
        "latex",
        caption="Results of PGLS analysis of ovariole number and egg size across a posterior distribution", booktabs = T, linesep = "") %>% 
    kableExtra::kable_styling(latex_options = "striped") %>%
    kableExtra::column_spec(1, width = "16em") %>%
    kableExtra::kable_styling(latex_options = "hold_position")  %>%
    kableExtra::kable_styling(font_size = 8) 
```

```{r PGLS-posterior-clade, include=TRUE}
### PGLS TABLE SUBCLADE
load("analysis/pgls_results_Fall2020/run_posterior_pgls_subclade.RData")

@@

clade_table <- as.data.frame(bind_rows(lapply(clade_posterior_results,get_pgls_table))) %>% mutate(analysis = clade_pgls_analysis_names,
        "taxonomic level" = clade_taxonomic_levels,
        "slope" = paste(sprintf("%.3f",slope_min),sprintf("%.3f",slope_max),sep=sp),
        "MCC p-value" = gsub("0.000","<0.001",sprintf("%.3f",sapply(clade_pgls_results,function(x){x$tTable[[8]]}))),
        "num. sig. / 1000" = pval_sig,
        "taxa" = taxa)
knitr::kable(clade_table %>% select(c("analysis","taxonomic level","slope","MCC p-value","num. sig. / 1000","taxa")),
        "latex",
        caption="Results of PGLS analysis of ovariole number and egg size across a posterior distribution", booktabs = T, linesep = "") %>% 
    kableExtra::kable_styling(latex_options = "striped") %>%
    kableExtra::column_spec(1, width = "16em") %>%
    kableExtra::kable_styling(latex_options = "hold_position")  %>%
    kableExtra::kable_styling(font_size = 8) 
```

```{r PGLS-posterior-body, include=TRUE}
### PGLS TABLE BODY SIZE
load("analysis/pgls_results_Fall2020/run_posterior_pgls_body_size.RData")

@@

body_pgls_analysis_names <- c("ovariole number vs body length",
    "ovariole number vs body weight",
    "ovariole number vs body weight",
    "Drosophilidae ovariole number vs thorax length",
    "Orthoptera ovariole number vs body length",
    "Curculionoidea ovariole number vs elytra length",
    "Hymenoptera ovariole number vs mesosoma width")

@@

body_table <- as.data.frame(bind_rows(lapply(body_posterior_results,get_pgls_table))) %>% mutate(analysis = body_pgls_analysis_names,
        "taxonomic level" = body_taxonomic_levels,
        "slope" = paste(sprintf("%.3f",slope_min),sprintf("%.3f",slope_max),sep=sp),
        "MCC p-value" = gsub("0.000","<0.001",sprintf("%.3f",sapply(body_pgls_results,function(x){x$tTable[[8]]}))),
        "num. sig. / 1000" = pval_sig,
        "taxa" = taxa)
knitr::kable(body_table %>% select(c("analysis","taxonomic level","slope","MCC p-value","num. sig. / 1000","taxa")),
        "latex",
        caption="Results of PGLS analysis of ovariole number and body size across a posterior distribution", booktabs = T, linesep = "") %>% 
    kableExtra::kable_styling(latex_options = "striped") %>%
    kableExtra::column_spec(1, width = "16em") %>%
    kableExtra::kable_styling(latex_options = "hold_position")  %>%
    kableExtra::kable_styling(font_size = 8) 
```

```{r PGLS-posterior-fecundity, include=TRUE}
### PGLS TABLE FECUNDITY
load("analysis/pgls_results_Fall2020/run_posterior_pgls_fecundity.RData")

@@

Using these 106 records for ovary type, we reconstructed the ancestral state at each node of the published phylogeny of `r length(pp$phy$tip.label)` insect genera using an equal-rates model that allows for missing data (Fig. \ref{fig:reconstruct-ovary}, R package corHMM,[@beaulieu2013identifying] version `r packageVersion('corHMM')`, function rayDISC, node.states = 'marginal').

@@

Using the ancestral state reconstruction of evolutionary shifts in the mode of oogenesis, we inferred the most likely mode of oogenesis for all nodes and unobserved extant tips in the phylogeny (Fig. \ref{fig:reconstruct-ovary}). We then compared the fit of models of trait evolution that take into account these shifts in oogenesis mode against those that do not. These comparisons were performed with the R package OUwie[@beaulieu2012modeling]  (version `r packageVersion('OUwie')`).

@@

We evaluated the fit of a Brownian Motion (BM) model for ovariole number evolution using the R package arbutus[@pennell2015model], version `r packageVersion('arbutus')`  (Fig. \ref{fig:arbutus}). In this approach, a BM model is fit to the data (R package geiger[@harmon2007geiger], version `r packageVersion('geiger')`), and the resulting parameters of the model are used to simulate 1000 new datasets. Six statistical parameters are used to compare the phylogenetic contrasts of the observed data to the simulated data, and their interpretations are as follows[@pennell2015model]:

@@

```{r BAMM-analysis-setup}
### Select a continuous trait to analyze
trait_of_interest <- "on"
dat_of_interest <- ON

@@

Given the result that our dataset contains substantial rate heterogeneity, we identified regions of the tree with high and low rates of ovariole number evolution using the software BAMM[@rabosky2014automatic], version 2.5.0. For this analysis, we calculated the average ovariole number for each genus in the insect phylogeny[@church2019insect]. Average ovariole number was log~10~ transformed, and the tree was filtered to include only tips for which there were corresponding ovariole number data (sample size = `r length(bmm$tip.label)`). We used the R package BAMMtools (version `r packageVersion('BAMMtools')`)[@rabosky2014bamm] to select priors, and ran BAMM for the maximum number of generations ($2*10^{9}$), sampling every 10^6^ generations.

```{r BAMM-analysis-convergence, include=TRUE, fig.cap="Convergence of trait diversification rate analysis", fig.width=10, fig.height=5}
### Assess convergence
mcmcout <- read.csv("analysis/BAMM_results_June2020/bamm_mcmc_out.txt", header=T)

@@

Convergence was evaluated both visually (Fig. \ref{fig:BAMM-analysis-convergence}) and numerically by comparing the effective sample size for number of shifts and log-likelihood to the standard recommended by the software (>200). We determined the most appropriate burn-in proportion to use by finding the maximum effective sample size of the log-likelihood across an array of possible burn-in proportions (Fig. \ref{fig:BAMM-analysis-burnin}). Running BAMM for the maximum possible number of generations and selecting the optimum burn-in\ref{fig:BAMM-analysis-burnin} resulted in an effective size for the number of shifts of `r round(es_nshifts,2)`, and for log-likelihood of `r round(es_logl,2)`. Repeated BAMM analyses showed similar distributions of high and low rate regimes, indicating the implications for ovariole number evolution are robust to uncertainty in rate estimates. See Supplemental Methods Section \ref{sec:BAMM} for details.

```{r BAMM-anlaysis-event-data}
### Read in the BAMM results and plot the shifts on the tree
bamm_tree <- ladderize(read.tree(file="analysis/BAMM_results_June2020/bamm_tree.tre"))

# get event data
edata <- getEventData(bamm_tree, eventdata = paste("analysis/BAMM_results_June2020/bamm_event_data.txt",sep=""), burnin=burnin, type="trait")
```

@@

The best configuration of rate shift regimes shows multiple independent clades with very low rates of evolution (Fig. \ref{fig:BAMM-analysis-results}). Visualizing the distribution of mean rates along branches revealed a discontinuous distribution, with one peak at a moderate rate of evolution and another at an extremely low rate, separated from the first peak by over six orders of magnitude (Fig. \ref{fig:BAMM-anlaysis-mean-rates}). We used this visualization to establish a threshold (`r lowrate`) for assigning a binary rate regime to each node in the phylogeny, categorizing them as above (variable) or below (invariant) a threshold that separates these two peaks.

@@

```{r BAMM-egg-setup}
tree <- genus_mcc_tree
trait_of_interest <- "vol"
dat_of_interest <- egg_database
dat <- dat_of_interest[sample(nrow(dat_of_interest)),] %>% # shuffle the dataset
mutate(raw_trait = eval(parse(text = trait_of_interest))) %>%  # pull out genus and trait
filter(!(is.na(raw_trait))) %>% select(genus,raw_trait) %>% # remove missing data
group_by(genus) %>% summarize(mean_trait = mean(raw_trait)) %>% # calculate mean per genus
mutate(trait = log10(mean_trait)) # transform the data

@@

```{r BAMM-egg-read}
egg_mcmcout <- read.csv("analysis/BAMM_results_June2020/egg_bamm_mcmc_out.txt", header=T)

>>>>

---
title: Repeated loss of variation in insect ovary morphology highlights the role of developmental constraint in life-history evolution - Supplementary Methods
author: 
    - Samuel H. Church^1^\*, 
    - Bruno A. S. De Medeiros^1,2^, 
    - Seth Donoughe^1,3^, 
    - Nicole L. Márquez Reyes^4^, 
    - Cassandra G. Extavour^1,5^\*
output:
    bookdown::pdf_document2:
    keep_tex: yes
    latex_engine: pdflatex
    #bookdown::html_document2: default
toc: true
bibliography: manuscript_supplement.bib
csl: nature.csl
link-citations: yes
header-includes:
 \usepackage{float}
 \usepackage{array}
 \usepackage{xcolor,colortbl}
 \floatplacement{figure}{H}
---

@@

```{r load_data, cache=FALSE}
# This script reads in and formats the ovariole number dataset
source("analysis/get_ovariole_number_data.R")
```

@@

Ovariole number was recorded as either an average with deviations, a range, or a single total value. When multiple types of data were available from a single publication, we recorded only a single type, with priority given to averages over ranges, and to both over single total values. Ovariole number was recorded as the total number of ovarioles per female, summing over both the left and right adult ovaries. When authors reported ovariole number from a single ovary, the total value was calculated by doubling the reported value. When authors described differences between the two ovaries, this information was recorded in an additional notes column.

@@

We combined the data we collected on total ovariole number with existing datasets of egg size and shape[@church2019dataset], insect lifetime fecundity and dry adult body mass[@Gilbert2007], average adult body length per insect family[@rainford2016phylogenetic], and several lineage-specific measures of adult body size[@iwata1966large;@Iwata1966a;@waloff1954number;@starmer2003phylogenetic;@reinhardt2005low].

Ovariole number and egg size[@church2019insect] data were combined by matching records at the species level (Fig 2a). When multiple records existed for a given species, the dataset was randomly shuffled and a single matching record was selected. This variation across records for the same species was accounted for in regressions by reshuffling and matching records at each iteration of the analysis.  We also matched records at the genus level following the same reshuffling method, which allowed us to test whether results were robust with a larger sample size when an exact species match was not available (Fig. \ref{fig:PGLS-egg-ON-genus}).

@@

Ovariole number, egg volume, lifetime fecundity, and adult body mass[@Gilbert2007] were combined by matching records at the species level and genus level, using the same method as described above (Figs. 2b, 3, and \ref{fig:PGLS-Gilbert-name}). We excluded one value from this dataset which appeared to include a typographical error for lifetime fecundity (Hymenoptera: Trichogrammatidae, lifetime fecundity recorded as 0.1[@Gilbert2007]).

@@

set.seed(12345)  # set seed for combination
source("analysis/combine_body_egg_datasets.R")

@@

```{r PGLS-Gilbert-name, fig.cap="**Egg volume vs ovariole number, phylogenetic residuals to dry adult body mass, matching records at the species level.**", fig.width=3, fig.height=3, include=TRUE}
### Gilbert egg vs ON, ind body weight name pgls
egg_ON_gilbert_name_pgls <- egg_ON_gilbert_name %>% 
    select(genus,logvol,logon,logweight,group) %>% 
    mutate(rank = genus, trait1 = logon, trait2= logvol, indep = logweight)

@@

We ran each Phylogenetic Generalized Least Squares (PGLS) regression over the Maximum Clade Credibilty (MCC) tree. For each regression, we also repeated each PGLS analysis 1000 times, accounting for phylogenetic and phenotypic uncertainty, using the R packages ape (version `r packageVersion('ape')`)[@paradis2004ape] and nlme (version `r packageVersion('nlme')`)[@pinheiro2014r]. In these analyses we used a Brownian Motion based covariance matrix for traits.

@@

For regressions of egg size and ovariole number when accounting for adult body size, we compared the results of our regression analyses to distributions estimated using simulated data under alternative hypotheses. We fit a Brownian motion model to the phylogenetic residuals of egg size and body size (R package geiger, version `r packageVersion('geiger')`)[@harmon2007geiger], and then used the parameters of this fitted model to simulate new datasets (R package phylolm, version `r packageVersion('phylolm')`)[@tung2014linear]. We performed this resimulation using the datasets of egg size and body length at the family level, and egg size and body mass at the genus level. We simulated 1000 datasets each under two hypotheses: no correlation (slope=0) and a strong negative correlation (slope=-1). We performed the regressions as described above and compared the distribution of p-values and slopes to values from regressions on observed data.

```{r simulated, fig.cap="**Using simulated data to test alternative hypotheses of evolutionary relationships.** Top row, distributions of p-values over 1000 replicate regressions, dashed black line indicates threshold of 0.01. Bottom row, distribution of estimated slopes between egg size and ovariole number, dashed lines indicate 95% interval of simulated distributions. Left, comparing egg size and ovariole number, accounting for body length at the family level. Right, comparing egg size and ovariole number, accounting for body mass at the genus level. Red=observed values, gray=simulated with no correlation, black=simulated with a -1:1 correlation. n=1000 regressions.", fig.width=7.5, fig.height=5, include=TRUE}
get_tstat <- function(sum,n) {
    tstat <- as.numeric(unname(sum[[n]]$coefficients[[2]]))
    return(tstat)
}

@@

load("analysis/pgls_results_Jan2021/run_pgls_simulated_allometry_body_egg_ON_family_-1_image.RData")
sim_corr_tstat_dist <- sapply(allometry_distribution_raw,get_tstat,n=1)
sim_corr_pval_dist <- sapply(allometry_distribution_raw,get_pval,n=1)
load("analysis/pgls_results_Jan2021/run_pgls_simulated_allometry_body_egg_ON_family_0_image.RData")
sim_nocorr_tstat_dist <- sapply(allometry_distribution_raw,get_tstat,n=1)
sim_nocorr_pval_dist <- sapply(allometry_distribution_raw,get_pval,n=1)
load("analysis/pgls_results_Jan2021/run_posterior_pgls_egg_size_ovariole_number.RData")
obs_tstat_dist <- sapply(allometry_distribution_raw,get_tstat,n=3)
obs_pval_dist <- sapply(allometry_distribution_raw,get_pval,n=3)

@@

load("analysis/pgls_results_Jan2021/run_pgls_simulated_allometry_gilbert_genus_-1_image.RData")
sim_corr_tstat_dist <- sapply(allometry_distribution_raw,get_tstat,n=1)
sim_corr_pval_dist <- sapply(allometry_distribution_raw,get_pval,n=1)
load("analysis/pgls_results_Jan2021/run_pgls_simulated_allometry_gilbert_genus_0_image.RData")
sim_nocorr_tstat_dist <- sapply(allometry_distribution_raw,get_tstat,n=1)
sim_nocorr_pval_dist <- sapply(allometry_distribution_raw,get_pval,n=1)
load("analysis/pgls_results_Jan2021/run_posterior_pgls_egg_size_ovariole_number.RData")
obs_tstat_dist <- sapply(allometry_distribution_raw,get_tstat,n=5)
obs_pval_dist <- sapply(allometry_distribution_raw,get_pval,n=5)

@@

p2 <- ggplot(gilbert_genus_pval_dist,aes(x=pval,y=relationship,color=relationship)) + geom_jitter(size=0.2)  + 
geom_vline(xintercept=0.01,colour="black",linetype="dashed") + 
#geom_vline(xintercept=0.05,colour="black",linetype="longdash") + 
ggtitle("egg size vs ov. number, body mass") + 
xlab("p-value") +
#ylab("type") + 
xlim(c(0,1)) +
scale_color_manual(values = c("black","dark gray","red")) +
theme(legend.position="none",
    axis.title.y=element_blank(),
axis.text.y=element_blank(),
axis.ticks.y=element_blank())

@@

```{r PGLS-body-plot, fig.cap="**Adult body size vs ovariole number.** Adult body mass compared at the genus level (left), and average adult body length compared at the family level (right).", fig.width=6, fig.height=3, include=TRUE}
### body length vs ON, family level
ON_body_family_pgls <- body_egg_ON_full %>% 
    select(rank,logon,logbody,group) %>% 
    rename(trait1 = logon, trait2 = logbody)

@@

gilbert_body <- run_plotting_pgls(ON_body_gilbert_genus_pgls,tree,"body_gilbert") + xlab(bquote('ovariole number,'~log[10])) +
ylab(bquote('dry adult body mass (g),'~log[10]))

@@

### PGLS TABLE EGG SIZE OVARIOLE NUMBER
load("analysis/pgls_results_Jan2021/run_posterior_pgls_egg_size_ovariole_number.RData")

@@

egg_ON_pgls_analysis_names <- c("ovariole number vs egg volume",
    "ovariole number vs egg volume",
    "ovariole number vs egg volume, residuals to body mass",
    "ovariole number vs egg volume, residuals to body mass",
    "ovariole number vs egg volume, residuals to body length")

@@

egg_ON_table <- as.data.frame(bind_rows(lapply(egg_ON_posterior_results,get_pgls_table))) %>% mutate(analysis = egg_ON_pgls_analysis_names,
        "taxonomic level" = egg_ON_taxonomic_levels,
        "slope" = paste(sprintf("%.3f",slope_min),sprintf("%.3f",slope_max),sep=sp),
        "MCC p-value" = gsub("0.000","<0.001",sprintf("%.3f",sapply(egg_ON_pgls_results,function(x){x$tTable[[8]]}))),
        "num. sig. / 1000" = pval_sig,
        "taxa" = taxa)
knitr::kable(egg_ON_table %>% select(c("analysis","taxonomic level","slope","MCC p-value","num. sig. / 1000","taxa")),
        "latex",
        align=rep('l', 6),
        caption="Results of PGLS analysis of ovariole number and egg size across a posterior distribution", booktabs = T, linesep = "") %>% 
    kableExtra::kable_styling(latex_options = "striped") %>%
    kableExtra::column_spec(1, width = "16em") %>%
    kableExtra::kable_styling(latex_options = "hold_position")  %>%
    kableExtra::kable_styling(font_size = 8) 
```

```{r PGLS-posterior-clade, include=TRUE}
### PGLS TABLE SUBCLADE
load("analysis/pgls_results_Jan2021/run_posterior_pgls_subclade.RData")

@@

clade_table <- as.data.frame(bind_rows(lapply(clade_posterior_results,get_pgls_table))) %>% mutate(analysis = clade_pgls_analysis_names,
        "taxonomic level" = clade_taxonomic_levels,
        "slope" = paste(sprintf("%.3f",slope_min),sprintf("%.3f",slope_max),sep=sp),
        "MCC p-value" = gsub("0.000","<0.001",sprintf("%.3f",sapply(clade_pgls_results,function(x){x$tTable[[8]]}))),
        "num. sig. / 1000" = pval_sig,
        "taxa" = taxa)
knitr::kable(clade_table %>% select(c("analysis","taxonomic level","slope","MCC p-value","num. sig. / 1000","taxa")),
        "latex",
        align=rep('l', 6),
        caption="Results of PGLS analysis of ovariole number and egg size across a posterior distribution", booktabs = T, linesep = "") %>% 
    kableExtra::kable_styling(latex_options = "striped") %>%
    kableExtra::column_spec(1, width = "16em") %>%
    kableExtra::kable_styling(latex_options = "hold_position")  %>%
    kableExtra::kable_styling(font_size = 8) 
```

```{r PGLS-posterior-body, include=TRUE}
### PGLS TABLE BODY SIZE
load("analysis/pgls_results_Jan2021/run_posterior_pgls_body_size.RData")

@@

body_pgls_analysis_names <- c("ovariole number vs body length",
    "ovariole number vs body mass",
    "ovariole number vs body mass",
    "Drosophilidae ovariole number vs thorax length",
    "Orthoptera ovariole number vs body length",
    "Curculionoidea ovariole number vs elytra length",
    "Hymenoptera ovariole number vs mesosoma width")

@@

body_table <- as.data.frame(bind_rows(lapply(body_posterior_results,get_pgls_table))) %>% mutate(analysis = body_pgls_analysis_names,
        "taxonomic level" = body_taxonomic_levels,
        "slope" = paste(sprintf("%.3f",slope_min),sprintf("%.3f",slope_max),sep=sp),
        "MCC p-value" = gsub("0.000","<0.001",sprintf("%.3f",sapply(body_pgls_results,function(x){x$tTable[[8]]}))),
        "num. sig. / 1000" = pval_sig,
        "taxa" = taxa)
knitr::kable(body_table %>% select(c("analysis","taxonomic level","slope","MCC p-value","num. sig. / 1000","taxa")),
        "latex",
        align=rep('l', 6),
        caption="Results of PGLS analysis of ovariole number and body size across a posterior distribution", booktabs = T, linesep = "") %>% 
    kableExtra::kable_styling(latex_options = "striped") %>%
    kableExtra::column_spec(1, width = "16em") %>%
    kableExtra::kable_styling(latex_options = "hold_position")  %>%
    kableExtra::kable_styling(font_size = 8) 
```

```{r PGLS-posterior-fecundity, include=TRUE}
### PGLS TABLE FECUNDITY
load("analysis/pgls_results_Jan2021/run_posterior_pgls_fecundity.RData")

@@

Using these 106 records for ovary type, we reconstructed the ancestral state at each node of the published phylogeny of `r length(pp$phy$tip.label)` insect genera using an equal-rates model that allows for missing data (Fig. \ref{fig:reconstruct-ovary}, R package corHMM, version 1.22[@beaulieu2013identifying] function rayDISC, node.states = 'marginal').

@@

Using the ancestral state reconstruction of evolutionary shifts in the mode of oogenesis, we inferred the most likely mode of oogenesis for all nodes and unobserved extant tips in the phylogeny (Fig. \ref{fig:reconstruct-ovary}). We then compared the fit of models of trait evolution that take into account these shifts in oogenesis mode against those that do not. These comparisons were performed with the R package OUwie (version 1.57)[@beaulieu2012modeling].

@@

We evaluated the fit of a Brownian Motion (BM) model for ovariole number evolution using the R package arbutus (version `r packageVersion('arbutus')`[@pennell2015model], Fig. \ref{fig:arbutus}). In this approach, a BM model is fit to the data (R package geiger, version `r packageVersion('geiger')`)[@harmon2007geiger], and the resulting parameters of the model are used to simulate 1000 new datasets. Six statistical parameters are used to compare the phylogenetic contrasts of the observed data to the simulated data, and their interpretations are as follows[@pennell2015model]:

@@

```{r BAMM-analysis-setup}
#set.seed(12345)  # set seed for BAMM

### Select a continuous trait to analyze
trait_of_interest <- "on"
dat_of_interest <- ON

@@

Given the result that our dataset contains substantial rate heterogeneity, we identified regions of the tree with high and low rates of ovariole number evolution using the software BAMM (version 2.5.0)[@rabosky2014automatic]. For this analysis, we calculated the average ovariole number for each genus in the insect phylogeny[@church2019insect]. Average ovariole number was log~10~ transformed, and the tree was filtered to include only tips for which there were corresponding ovariole number data (sample size = `r length(bmm$tip.label)`). We used the R package BAMMtools (version `r packageVersion('BAMMtools')`)[@rabosky2014bamm] to select priors, and ran BAMM for the maximum number of generations ($2*10^{9}$), sampling every 10^6^ generations.

```{r BAMM-analysis-convergence, include=TRUE, fig.cap="Convergence of trait diversification rate analysis", fig.width=10, fig.height=5}
### Assess convergence
mcmcout <- read.csv("analysis/BAMM_results_Jan2021/bamm_mcmc_out.txt", header=T)

@@

Convergence was evaluated both visually (Fig. \ref{fig:BAMM-analysis-convergence}) and numerically by comparing the effective sample size for number of shifts and log-likelihood to the standard recommended by the software (>200). We determined the most appropriate burn-in proportion to use by finding the maximum effective sample size of the log-likelihood across an array of possible burn-in proportions (Fig. \ref{fig:BAMM-analysis-burnin}). Running BAMM for the maximum possible number of generations and selecting the optimum burn-in (Fig. \ref{fig:BAMM-analysis-burnin}) resulted in an effective size for the number of shifts of `r round(es_nshifts,2)`, and for log-likelihood of `r round(es_logl,2)`. Repeated BAMM analyses showed similar distributions of high and low rate regimes, indicating the implications for ovariole number evolution are robust to uncertainty in rate estimates. See Supplemental Methods Section \ref{sec:BAMM} for details.

```{r BAMM-anlaysis-event-data}
### Read in the BAMM results and plot the shifts on the tree
bamm_tree <- ladderize(read.tree(file="analysis/BAMM_results_Jan2021/bamm_tree.tre"))

# get event data
edata <- getEventData(bamm_tree, eventdata = paste("analysis/BAMM_results_Jan2021/bamm_event_data.txt",sep=""), burnin=burnin, type="trait")
```

@@

The best configuration of rate shift regimes shows multiple independent clades with very low rates of evolution (Fig. \ref{fig:BAMM-analysis-results}). Visualizing the distribution of mean rates along branches revealed a discontinuous distribution, with one peak at a moderate rate of evolution and several clusters at extremely low rates, separated from the first peak by over six orders of magnitude (Fig. \ref{fig:BAMM-anlaysis-mean-rates}). We used this visualization to establish a threshold (`r lowrate`) for assigning a binary rate regime to each node in the phylogeny, categorizing them as above (variable) or below (invariant) a threshold that separates these two peaks.

@@

```{r BAMM-egg-setup}
#set.seed(12345)  # set seed for egg BAMM

tree <- genus_mcc_tree
trait_of_interest <- "vol"
dat_of_interest <- egg_database
dat <- dat_of_interest[sample(nrow(dat_of_interest)),] %>% # shuffle the dataset
    mutate(raw_trait = eval(parse(text = trait_of_interest))) %>%  # pull out genus and trait
    filter(!(is.na(raw_trait))) %>% select(genus,raw_trait) %>% # remove missing data
    group_by(genus) %>% summarize(mean_trait = mean(raw_trait)) %>% # calculate mean per genus
    mutate(trait = log10(mean_trait)) # transform the data

@@

```{r BAMM-egg-read}
egg_mcmcout <- read.csv("analysis/BAMM_results_Jan2021/egg_bamm_mcmc_out.txt", header=T)
